---
title: "MFRE R Prep: Assignment"
author: "Your Name Here (replace with your full name)"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Instructions

Complete this assignment after working through `r_intro.Rmd` and `r_applications.Rmd`.

- Replace the author field at the top of this file with your full name.
- Write your answer in the code cell under each question.
- Add at least one comment per code cell explaining your reasoning.
- Run each cell as you go (`Ctrl + Enter`, or `Cmd + Enter` on macOS).
- When finished, click **Knit** to confirm the whole file runs without errors.
- Email both the `.Rmd` file and the knitted `.html` file to `olivier.ntwali@ubc.ca`.
- Subject line: `MFRE R Prep - Assignment - [Your Name]`

The assignment uses base R only. No extra packages are needed.

# Dataset

Run this cell first. All questions use these objects or the built-in `iris` and `mtcars` datasets.

```{r}
# Crop yield data (tonnes per hectare, approximate 2023 values)
countries  <- c("Canada", "USA", "Australia", "Russia", "France")
crops      <- c("wheat", "corn", "wheat", "wheat", "wheat")
yields_tha <- c(3.50, 9.80, 2.10, 3.00, 7.20)
exports_mt <- c(22.0, 26.5, 21.0, 33.0, 18.0)

crop_data <- data.frame(
  country    = countries,
  crop       = crops,
  yield_t_ha = yields_tha,
  export_mt  = exports_mt
)

# Conversion factors: 1 t/ha to bushels/acre
bushel_factors <- c(wheat = 14.87, corn = 39.37, canola = 44.09, soy = 14.87)

print(crop_data)
```

# Question 1: Vectors and arithmetic

a) Calculate and print the average yield across all five rows of `yields_tha`.

b) France has the highest wheat yield in the dataset and Australia the lowest. Calculate how many times higher France's yield is than Australia's. Print the result rounded to 2 decimal places.

c) Convert Canada's wheat yield from t/ha to bushels per acre using `bushel_factors`. Print the result rounded to 1 decimal place.

```{r}
# Your code here

```

# Question 2: Control flow

Write a `for` loop over `yields_tha` that prints one line per country in this format:

```
Canada: 3.5 t/ha (Good)
```

The rating in parentheses follows this table. Use `if`, `else if`, and `else` inside the loop, and `sprintf()` to build each line.

| Yield (t/ha)  | Rating    |
|---------------|-----------|
| 6.0 or higher | Excellent |
| 3.0 to 6.0    | Good      |
| 1.5 to 3.0    | Fair      |
| below 1.5     | Poor      |

```{r}
# Your code here

```

# Question 3: Writing a function

Write a function `convert_yield(yield_t_ha, crop)` that:

1. Looks up the conversion factor for `crop` in `bushel_factors`.
2. Returns the string `"Error: no conversion factor for <crop>"` if the crop is not in `bushel_factors`. The `%in%` operator and `names()` are useful here.
3. Otherwise returns the yield in bushels per acre, rounded to 1 decimal place.

Test it with:

```r
convert_yield(3.50, "wheat")   # 52.0
convert_yield(9.80, "corn")    # 385.8
convert_yield(3.00, "rice")    # error message
```

```{r}
# Your code here

```

# Question 4: Data frames

Using `crop_data`:

a) Filter the rows where `yield_t_ha` is greater than 3.0 and print them.

b) Add a column `yield_kg_ha` that converts the yield to kg per hectare (multiply by 1000). Print the updated data frame.

c) Calculate the average `export_mt` of the wheat rows only.

```{r}
# Your code here

```

# Question 5: Aggregation and plotting

a) Using the built-in `iris` dataset, calculate the mean `Sepal.Length` for each `Species` with `aggregate()`.

b) Using `crop_data`, create a scatter plot of `export_mt` against `yield_t_ha`. Label both axes and add a title.

```{r}
# Your code here

```

# Question 6: Correlation and regression

Using the built-in `mtcars` dataset:

a) Calculate the correlation between `wt` (weight) and `mpg`.

b) Fit a linear regression predicting `mpg` from `wt` and print the model summary.

c) In a comment, state whether the relationship is positive or negative and what that means in one sentence.

```{r}
# Your code here

```

# Submission checklist

- [ ] Your name is in the author field at the top of this file
- [ ] Every question has code in its answer cell, with comments
- [ ] The file knits to HTML without errors
- [ ] You emailed both the `.Rmd` and the `.html` to `olivier.ntwali@ubc.ca`
