---
title: "MFRE R Prep: Applications"
author: "MFRE Program, University of British Columbia"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Install required packages only if they are missing, then load them.
required <- c("zoo", "class", "ggplot2")
missing_pkgs <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_pkgs)) {
  install.packages(missing_pkgs, repos = "https://cloud.r-project.org")
}
library(zoo)
library(class)
library(ggplot2)
```

# Introduction

Work through the Python notebooks and the R intro before this one. Here we apply the basics to small, real problems: data manipulation, cleaning, visualisation, analysis, and a short machine learning example.

> **Before you start.** Install R and RStudio first (see `Installation Guide/README_setup.md`). Open this file in RStudio and click **Knit**, or run cells with `Ctrl + Enter`. The setup chunk below installs the `zoo`, `class`, and `ggplot2` packages if they are missing, so the first run needs an internet connection. You can also install them yourself once with `install.packages(c("zoo", "class", "ggplot2"))`.

# Data Manipulation

R ships with built-in datasets. We use `mtcars`, which holds specifications for several car models. Run `?mtcars` in the console for the full description.

```{r}
head(mtcars)
```

Load it into a data frame and filter it.

```{r}
df <- mtcars

# Keep cars with mpg greater than 20
df_filtered <- df[df$mpg > 20, ]
head(df_filtered)

# Average horsepower of those cars
mean_hp <- mean(df_filtered$hp)
print(mean_hp)
```

Often you need to extract a subset. `mtcars` has no manufacturer column, but the row names are in the form `manufacturer model`, so we can split on the space.

```{r}
# Take the first word of each row name as the manufacturer
df$manufacturer <- sapply(strsplit(rownames(df), " "), "[", 1)

# Average horsepower by manufacturer
df_summary <- aggregate(hp ~ manufacturer, data = df, FUN = mean)
print(df_summary)
```

How this works:

1. `strsplit` splits each row name on the space and returns a list.
2. `sapply(..., "[", 1)` takes the first element of each split, the manufacturer.
3. `aggregate` averages `hp` within each manufacturer.

# Data Cleaning

Cleaning means handling missing values and errors before analysis. We use `iris` and deliberately introduce problems.

The `Species` column is a factor, so we first convert it to character. Without this step, assigning a new label like `"error"` would produce `NA` and a warning instead of the text we want.

```{r}
df <- iris
df$Species <- as.character(df$Species)   # convert factor so we can inject text errors

set.seed(42)   # fix the random draw so results are reproducible

# Introduce missing values and errors
df[sample(seq_len(nrow(df)), 10), "Sepal.Length"] <- NA
df[sample(seq_len(nrow(df)), 10), "Species"] <- "error"

print(sum(is.na(df)))                  # count missing values
print(sum(df$Species == "error"))      # count injected errors
```

Now clean it.

```{r}
# Drop rows with missing values
df_clean <- df[complete.cases(df), ]

# Drop rows with the error label
df_clean <- df_clean[df_clean$Species != "error", ]

print(sum(is.na(df_clean)))               # 0
print(sum(df_clean$Species == "error"))   # 0
print(dim(df_clean))                       # rows remaining
```

`complete.cases` returns `TRUE` for rows with no missing values, which we use to subset. We then drop rows whose `Species` is `"error"`.

For time series with gaps, you can carry the last known value forward with `na.locf` from the `zoo` package.

```{r}
ts_data <- zoo(
  c(1, NA, 3, NA, 5),
  as.Date(c("2022-01-01", "2022-01-02", "2022-01-03", "2022-01-04", "2022-01-05"))
)
print(ts_data)

ts_filled <- na.locf(ts_data)   # last observation carried forward
print(ts_filled)
```

# Data Visualisation

Base R covers the common plot types.

```{r}
df <- mtcars

# Scatter plot
plot(df$hp, df$mpg,
     xlab = "Horsepower", ylab = "Miles per Gallon",
     main = "MPG vs HP")
```

```{r}
# Boxplot of mpg by number of cylinders
boxplot(df$mpg ~ df$cyl,
        xlab = "Number of Cylinders", ylab = "Miles per Gallon",
        main = "MPG by Cylinders")
```

```{r}
# Histogram of mpg
hist(df$mpg, xlab = "Miles per Gallon", main = "Distribution of MPG")
```

For more control and interactivity, use packages such as `ggplot2` or `plotly`.

# Data Analysis

Quantify the relationship between two variables with correlation.

```{r}
df <- mtcars
correlation <- cor(df$mpg, df$hp)
print(correlation)
```

Correlation ranges from -1 to 1: near 1 is a strong positive relationship, near -1 a strong negative one, near 0 no linear relationship.

Fit a linear regression to predict `mpg` from `hp`.

```{r}
model <- lm(mpg ~ hp, data = df)
summary(model)
```

`lm` fits the model and `summary` reports the coefficients, standard errors, t-values, and p-values.

# Machine Learning Example

A short k-nearest neighbours (KNN) example. KNN classifies a point by the majority class of its `k` nearest neighbours. We split `iris`, train on the sepal measurements, and predict species.

```{r}
df <- iris

set.seed(123)
train_indices <- sample(seq_len(nrow(df)), 0.7 * nrow(df))
train_data <- df[train_indices, ]
test_data  <- df[-train_indices, ]

predicted <- knn(
  train = train_data[, c("Sepal.Length", "Sepal.Width")],
  test  = test_data[, c("Sepal.Length", "Sepal.Width")],
  cl    = train_data$Species,
  k     = 3
)

# Confusion matrix: predicted vs actual
confusion_matrix <- table(Predicted = predicted, Actual = test_data$Species)
print(confusion_matrix)
```

The confusion matrix compares predictions to the true species. From it you can compute accuracy and other metrics.

We can also visualise the predictions. Here we train on all four measurements and plot the predicted species by petal size.

```{r}
set.seed(123)
ind <- sample(2, nrow(iris), replace = TRUE, prob = c(0.7, 0.3))
iris_train <- iris[ind == 1, 1:4]
iris_test  <- iris[ind == 2, 1:4]
train_labels <- iris[ind == 1, 5]

knn_pred <- knn(iris_train, iris_test, train_labels, k = 3)

plot_df <- data.frame(iris_test, Species = knn_pred)

ggplot(plot_df, aes(Petal.Length, Petal.Width, color = Species)) +
  geom_point() +
  labs(title = "KNN classification of iris",
       x = "Petal Length", y = "Petal Width")
```

Predict on new measurements.

```{r}
new_data <- data.frame(
  Sepal.Length = c(5.1, 6.2, 7.3),
  Sepal.Width  = c(3.5, 2.9, 2.8)
)

predictions <- knn(
  train = train_data[, c("Sepal.Length", "Sepal.Width")],
  test  = new_data,
  cl    = train_data$Species,
  k     = 3
)
print(predictions)
```

For larger problems, packages such as `caret`, `randomForest`, and `xgboost` provide more models and tooling.

# Exercises

Use the `mtcars` dataset. Attempt each question before revealing the solution. Solutions are commented out at the bottom of each cell.

```{r}
head(mtcars)
```

**1.** Filter the cars with `mpg` greater than 20 and `hp` greater than 100. What is the average `wt` of these cars?

```{r}
# Your code here


# --- Solution (uncomment to check) ---
# subset_cars <- mtcars[mtcars$mpg > 20 & mtcars$hp > 100, ]
# mean(subset_cars$wt)
```

**2.** Add a column `disp_per_cyl` to `mtcars` that is `disp` divided by `cyl`.

```{r}
# Your code here


# --- Solution (uncomment to check) ---
# mtcars$disp_per_cyl <- mtcars$disp / mtcars$cyl
# head(mtcars)
```

**3.** Create a scatter plot of `disp_per_cyl` against `mpg`, with axis labels and a title.

```{r}
# Your code here


# --- Solution (uncomment to check) ---
# mtcars$disp_per_cyl <- mtcars$disp / mtcars$cyl
# plot(mtcars$disp_per_cyl, mtcars$mpg,
#      xlab = "Displacement per Cylinder", ylab = "Miles per Gallon",
#      main = "MPG vs Displacement per Cylinder")
```

**4.** Create a boxplot of `disp_per_cyl` by `cyl`, with axis labels and a title.

```{r}
# Your code here


# --- Solution (uncomment to check) ---
# mtcars$disp_per_cyl <- mtcars$disp / mtcars$cyl
# boxplot(mtcars$disp_per_cyl ~ mtcars$cyl,
#         xlab = "Number of Cylinders", ylab = "Displacement per Cylinder",
#         main = "Displacement per Cylinder by Cylinders")
```

**5.** Calculate the correlation between `disp_per_cyl` and `mpg`.

```{r}
# Your code here


# --- Solution (uncomment to check) ---
# mtcars$disp_per_cyl <- mtcars$disp / mtcars$cyl
# cor(mtcars$disp_per_cyl, mtcars$mpg)
```

**6.** Fit a linear regression predicting `mpg` from `disp_per_cyl`, and print the model summary.

```{r}
# Your code here


# --- Solution (uncomment to check) ---
# mtcars$disp_per_cyl <- mtcars$disp / mtcars$cyl
# model <- lm(mpg ~ disp_per_cyl, data = mtcars)
# summary(model)
```
