---
title: "MFRE R Prep: Introduction to R"
author: "MFRE Program, University of British Columbia"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

Complete the Python prep section before starting here. Many ideas carry over from Python, but R has its own syntax and conventions. This notebook covers the R basics you need before the summer bootcamp.

R is a language for statistical computing and graphics. You will rarely use it for general programming, but it is a strong tool for data analysis and visualisation because of its large package ecosystem.

## R Markdown

We use R Markdown (`.Rmd`) files rather than plain `.R` scripts. An `.Rmd` file combines text, code, and output in one document that you can render ("knit") to HTML, PDF, or Word.

Like a Jupyter notebook, an `.Rmd` is made of cells. Run a code cell with the Run button or `Ctrl + Enter`.

> **Before you start.** Install R and RStudio first. Follow the Installation Guide in this package (`Installation Guide/README_setup.md`). Open this file in RStudio, then click **Knit** to render the whole document, or run one cell at a time with `Ctrl + Enter`. This notebook uses only base R, so no extra packages are needed.

```{r}
first_value <- 5
second_value <- 10
result <- first_value + second_value
print(result)
```

Note the first syntax difference from Python: R assigns with `<-`. The code above assigns 5 to `first_value`, 10 to `second_value`, adds them, and prints the result.

## Functional vs object-oriented style

R leans functional: you mostly apply functions to data. Python leans object-oriented, where you call methods on objects. In practice this means you will often write `function(data)` in R where you would write `data.method()` in Python.

```python
# Python: method called on the list object
array_1 = [1, 2, 3, 4, 5]
array_1.append(6)
print(array_1)
```

```{r}
# R: combine with a function instead of a method
array_1 <- c(1, 2, 3, 4, 5)
array_1 <- c(array_1, 6)
print(array_1)

# Comments in R use the hash symbol, same as Python
```

**This notebook covers:**

1. Data types
2. Data structures
3. Basic operations
4. Control flow
5. Functions
6. Variable naming
7. Pseudocode
8. Debugging
9. Resources

# Data Types

The common R data types are:

1. **Numeric**: numbers, including integers and decimals (floats).
2. **Character**: text strings.
3. **Logical**: `TRUE` or `FALSE`.
4. **Integer**: whole numbers (created with `L` or `as.integer()`).
5. **Complex**: complex numbers.
6. **Raw**: raw bytes.

```{r}
# Numeric
num_value <- 5
print(num_value)

# Character
char_value <- "Hello, World!"
print(char_value)

# Logical
log_value <- TRUE
print(log_value)

# Integer
int_value <- as.integer(5)
print(int_value)

# Complex
cplx_value <- 5 + 3i
print(cplx_value)

# Raw
raw_value <- charToRaw("Hello, World!")
print(raw_value)
```

Be aware of types when you mix them. Some combinations error, others coerce silently.

```{r}
# Numeric plus character: this errors, so we catch it
num_value <- 5
char_value <- "10"
tryCatch(
  print(num_value + char_value),
  error = function(e) print(e)
)

# Numeric plus logical: TRUE counts as 1, so this returns 6
num_value <- 5
log_value <- TRUE
print(num_value + log_value)
```

# Data Structures

The main R data structures are:

1. **Vectors**: one-dimensional, all elements the same type.
2. **Matrices**: two-dimensional, all elements the same type.
3. **Lists**: one-dimensional, mixed types allowed.
4. **Data frames**: two-dimensional, mixed types by column. This is the closest analogue to a pandas DataFrame.

```{r}
# Vector
vec <- c(1, 2, 3, 4, 5)
print(vec)

# Matrix
mtx <- matrix(1:9, nrow = 3, ncol = 3)
print(mtx)

# List
li <- list(1, "Hello, World!", TRUE)
print(li)

# Data frame
df <- data.frame(x = c(1, 2, 3, 4, 5), y = c("a", "b", "c", "d", "e"))
print(df)
```

# Basic Operations

R supports the standard arithmetic operators.

```{r}
first_value <- 5
second_value <- 10

print(first_value + second_value)   # addition
print(first_value - second_value)   # subtraction
print(first_value * second_value)   # multiplication
print(first_value / second_value)   # division
print(first_value %% second_value)  # remainder (modulo)
print(first_value ^ second_value)   # exponent
```

Use `sprintf()` to build a formatted string.

```{r}
result <- first_value + second_value
print(sprintf("The sum of %d and %d is %d", first_value, second_value, result))
```

# Control Flow

R supports `if`/`else`, `for` loops, and `while` loops.

```{r}
# If-else
value <- 5
if (value > 0) {
  print("value is positive")
} else {
  print("value is not positive")
}

# For loop
for (i in 1:5) {
  print(i)
}

# While loop
i <- 1
while (i <= 5) {
  print(i)
  i <- i + 1
}
```

Boolean operators let you combine or negate conditions. Wrap the negated condition in parentheses so the comparison runs before the `!`.

```{r}
first_value <- 5
second_value <- 10

if (first_value > 0 & second_value > 0) {
  print("Both values are positive")
}

if (first_value > 0 | second_value > 0) {
  print("At least one value is positive")
}

if (!(first_value > 0)) {
  print("first_value is not positive")
} else {
  print("first_value is positive")
}
```

Use `%in%` to test whether a value is a member of a vector. To test whether a substring appears inside a string, use `grepl()`.

```{r}
# Membership in a vector
fruits <- c("apple", "banana", "cherry")
print("banana" %in% fruits)   # TRUE

# Substring inside a string
print(grepl("World", "Hello, World!"))   # TRUE
```

R uses curly braces to mark code blocks, unlike Python which uses indentation.

# Functions

Define a function with the `function` keyword. There is no `def` keyword as in Python.

```{r}
add <- function(first_value, second_value) {
  result <- first_value + second_value
  return(result)
}

print(add(5, 10))
```

A function can call another function. Here `double_value` calls `add`.

```{r}
add <- function(first_value, second_value) {
  return(first_value + second_value)
}

double_value <- function(value) {
  return(add(value, value))
}

print(double_value(5))   # 10
```

Define a function before you call it, otherwise R raises an error.

A recursive function calls itself. The classic example is factorial.

```{r}
factorial_recursive <- function(n) {
  if (n == 0) {
    return(1)            # base case
  } else {
    return(n * factorial_recursive(n - 1))   # recursive case
  }
}

print(factorial_recursive(5))   # 120
```

Recursion suits problems with a recursive structure, but it can be harder to debug than a loop and can overflow the stack if it goes too deep. Use it where it fits.

# Variable Naming

Variable names can contain letters, numbers, and underscores, and cannot start with a number. Names are case-sensitive, so `x`, `X`, and `X1` differ.

```{r, eval=FALSE}
# Valid
first_value <- 5
y <- "Hello, World!"
z_1 <- TRUE
my_variable <- 10

# Invalid (these would error)
1x <- 5
hello world <- "Hello, World!"
my-variable <- 10
```

Pick one naming convention and keep it consistent: `camelCase`, `snake_case`, or `PascalCase`. Use descriptive names (`total_sum`, not `x`), but keep them short enough to type.

# Pseudocode

Pseudocode is a plain-language sketch of an algorithm. It helps you plan the logic before writing real code, and it forms the basis of good comments.

```
# Sum of the first n natural numbers
set sum to 0
for i from 1 to n
    add i to sum
return sum
```

# Debugging

When you are stuck, a few habits help:

1. **Explain the code out loud**, line by line ("rubber duck debugging"). Stating each step often exposes the error.
2. **Break the problem down** and test small pieces separately.
3. **Print intermediate values** to see what the code is actually doing.
4. **Read the error message**: the last line usually names the problem.
5. **Take a break** and return with fresh eyes.

# Resources

1. R documentation: https://cran.r-project.org/manuals.html (or type `?function_name`).
2. *R for Data Science* by Wickham and Grolemund: https://r4ds.hadley.nz/
3. Stack Overflow for specific errors and questions.
4. RStudio Community: https://community.rstudio.com/

# Notes on R Markdown

An `.Rmd` file is rendered to HTML, PDF, or Word by the `knitr` package. The output format is set in the YAML header at the top of the file (this notebook uses `html_document`). A minimal header looks like this:

```{r, eval=FALSE}
---
title: "My R Markdown Document"
output: html_document
---
```

# Assignment

Attempt each question before revealing the solution. Solutions are commented out at the bottom of each cell. Uncomment to check your work.

**1.** Construct a 2 x 3 matrix holding the values 1 to 6 in descending order, filled row by row.

```{r}
# Your code here


# --- Solution (uncomment to check) ---
# matrix(6:1, nrow = 2, ncol = 3, byrow = TRUE)
```

**2.** Write a function that prints a pyramid of `*` characters. It takes a positive integer and prints that many levels.

```{r}
# Your code here


# --- Solution (uncomment to check) ---
# print_pyramid <- function(n) {
#   for (i in 1:n) {
#     spaces <- strrep(" ", n - i)
#     stars  <- strrep("*", 2 * i - 1)
#     cat(spaces, stars, "\n", sep = "")
#   }
# }
# print_pyramid(5)
```

**3.** The function below should return the cumulative sum of the numbers from 1 to `n`, but it has a bug. Find and fix it.

```{r}
cum_sum <- function(n) {
  total <- 0
  for (i in 1:n) {
    total <- i          # bug: this overwrites total instead of adding to it
  }
  return(total)
}

print(cum_sum(5))   # returns 5, but should return 15

# --- Solution (uncomment to check) ---
# cum_sum <- function(n) {
#   total <- 0
#   for (i in 1:n) {
#     total <- total + i
#   }
#   return(total)
# }
# print(cum_sum(5))   # 15
```

**4.** You are given a 16-digit credit card number as a string. Write a function that returns `TRUE` if the sum of its digits is divisible by 10, and `FALSE` otherwise.

```{r}
card_number <- "1234567890123456"

# Your code here


# --- Solution (uncomment to check) ---
# is_valid_card <- function(card) {
#   digits <- as.integer(strsplit(card, "")[[1]])
#   sum(digits) %% 10 == 0
# }
# is_valid_card(card_number)   # FALSE (digit sum is 66)
```

**5.** The data frame below has cities as rows and dates as columns. Add a logical column `any_above_30` that is `TRUE` if the city recorded a temperature above 30 on any date.

```{r}
df <- data.frame(
  city   = c("City A", "City B", "City C"),
  date_1 = c(25, 30, 35),
  date_2 = c(28, 29, 30),
  date_3 = c(30, 32, 33)
)

# Your code here


# --- Solution (uncomment to check) ---
# df$any_above_30 <- apply(df[, c("date_1", "date_2", "date_3")], 1,
#                          function(row) any(row > 30))
# print(df)
```
